The Design and Implementation of the Synchronous Language \CEU

### 2016-04-01 paper.tex
...

@@

@@

% unique external events

The notion of time based on events instead of ticks 
(Section~\ref{sec.ceu.exts}) is widespread~\cite{sync_async.whynotthreads} and 
popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
Like \CEU, event-driven programming is essentially synchronous, i.e., events go 
through a queue and are dispatched sequentially and atomically to prevent race 
conditions.
%
We believe that for software design, this approach is more familiar to 
programmers and simplifies the reasoning about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis for concurrent statements.

% internal events

Internal events

% lexical order

@@

lexical order pragamatics
    - again, familiar to programmers
        - register order
    - paralax, logging
        - simple, understandeable, imperative
        - multiple calls to compose, printf, are not commutative
        - define with intimidating name
    - implementation aspects in Section XX

>>>>

From our experience, however, we find that it is not uncommon to require 
stateful and non-commutative calls to execute concurrently.
%
This is the case for logging (e.g., calls to \code{\_printf} in parallel) or 
redrawing objects in the screen.
%
Figure~\ref{fig.redraw}.a shows an abstract code to animate and redraw objects 
\code{back} and \code{fore} in the screen in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels.
This way, \code{\_redraw} is non commutative and the code does not compile 
correctly.
%
However, since the \code{back} object should be redrawn before the \code{fore} 
object, we want to rely on \CEU's intra-reaction deterministic execution based 
on lexical order.
Therefore, in Figure~\ref{fig.redraw}, we redefine the redrawing function to an 
``intimidating'' name \code{\_redraw\_non\_commutative} to be explicit about 
its effect (ln. 2), and annotate it as safe (ln. 4--5) to compile successfully.

\begin{figure}
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}
par do
    <...> // animate and redraw "back"
        _redraw(back);
with
    <...> // animate and redraw "fore"
        _redraw(fore);
end

.
\end{lstlisting}
\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
\end{minipage}
%
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define redraw_non_commutative redraw
end
@safe _redraw_non_commutative with
      _redraw_non_commutative;
par do
    <...> // animate and redraw "back"
        _redraw_non_commutative(back);
with
    <...> // animate and redraw "fore"
        _redraw_non_commutative(fore);
end
\end{lstlisting}
\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
\label{lst.c}
}
\end{figure}

@@

\CEU is a synchronous language inspired by Esterel which aims to offer a 
simpler semantics and more fine-grained control over the program execution.
%
\CEU employs an event-triggered notion of time that allows for a simple
reasoning about reactions in parallel and enables compile-time analysis to 
produce deterministic and concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.

Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.

@@

@@

lexical order pragamatics
    - again, familiar to programmers
        - register order
    - implementation aspects in Section XX
